---
title: "Clustering Comparison: Standard vs IDEAL"
subtitle: "Analyzing Differences Between Clustering Approaches"
author: "Clustering Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 12
    fig_height: 8
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(knitr)

# Set working directory to project root
knitr::opts_knit$set(root.dir = '../../')
```

## Data Loading

```{r load_data}
# Load clustering data from both approaches
clustering_standard <- list(
  groups = read.csv('data/results/clustering/groups.csv', row.names = 1),
  chars = read.csv('data/results/clustering/characteristics.csv'),
  centroids = read.csv('data/results/clustering/centroids.csv'),
  kcca_info = read.csv('data/results/clustering/kcca_info.csv')
)

clustering_ideal <- list(
  groups = read.csv('data/results/clustering_IDEAL/groups.csv', row.names = 1),
  chars = read.csv('data/results/clustering_IDEAL/characteristics.csv'),
  centroids = read.csv('data/results/clustering_IDEAL/centroids.csv'),
  kcca_info = read.csv('data/results/clustering_IDEAL/kcca_info.csv')
)

# Load most important attributes data
mia_data <- read.csv('data/results/most_important_attr/mia.csv')

# Display basic info
cat("Standard Clustering - Number of datasets:", nrow(clustering_standard$groups), "\n")
cat("IDEAL Clustering - Number of datasets:", nrow(clustering_ideal$groups), "\n")
```

## Clustering Overview

### Standard Clustering Summary

```{r standard_summary}
# Summary statistics for standard clustering
cat("K-value:", clustering_standard$kcca_info$K[1], "\n")
cat("Number of clusters:", clustering_standard$kcca_info$num_centers[1], "\n")
cat("Total within-cluster sum of squares:", round(clustering_standard$kcca_info$tot_withinss[1], 4), "\n")
cat("Between-cluster sum of squares:", round(clustering_standard$kcca_info$betweenss[1], 4), "\n")
cat("Cluster sizes:", paste(clustering_standard$kcca_info[1, grepl("size_cluster", names(clustering_standard$kcca_info))], collapse = ", "), "\n")
```

### IDEAL Clustering Summary

```{r ideal_summary}
# Summary statistics for ideal clustering
cat("K-value:", clustering_ideal$kcca_info$K[1], "\n")
cat("Number of clusters:", clustering_ideal$kcca_info$num_centers[1], "\n")
cat("Total within-cluster sum of squares:", round(clustering_ideal$kcca_info$tot_withinss[1], 4), "\n")
cat("Between-cluster sum of squares:", round(clustering_ideal$kcca_info$betweenss[1], 4), "\n")
cat("Cluster sizes:", paste(clustering_ideal$kcca_info[1, grepl("size_cluster", names(clustering_ideal$kcca_info))], collapse = ", "), "\n")
```

## Group Assignment Comparison

```{r group_comparison}
# Create comparison dataframe
comparison_df <- data.frame(
  dataset_name = clustering_standard$groups$dataset_name,
  standard_hclust_4 = clustering_standard$groups$hclust_4,
  ideal_hclust_4 = clustering_ideal$groups$hclust_4,
  standard_hclust_5 = clustering_standard$groups$hclust_5,
  ideal_hclust_5 = clustering_ideal$groups$hclust_5,
  standard_kmeans_4 = clustering_standard$groups$kmeans_4,
  ideal_kmeans_4 = clustering_ideal$groups$kmeans_4,
  standard_kmeans_5 = clustering_standard$groups$kmeans_5,
  ideal_kmeans_5 = clustering_ideal$groups$kmeans_5
)

# Calculate changes
comparison_df$hclust_4_changed <- comparison_df$standard_hclust_4 != comparison_df$ideal_hclust_4
comparison_df$hclust_5_changed <- comparison_df$standard_hclust_5 != comparison_df$ideal_hclust_5
comparison_df$kmeans_4_changed <- comparison_df$standard_kmeans_4 != comparison_df$ideal_kmeans_4
comparison_df$kmeans_5_changed <- comparison_df$standard_kmeans_5 != comparison_df$ideal_kmeans_5

# Summary of changes
summary_changes <- data.frame(
  Algorithm = c("HClust K=4", "HClust K=5", "KMeans K=4", "KMeans K=5"),
  Datasets_Changed = c(
    sum(comparison_df$hclust_4_changed),
    sum(comparison_df$hclust_5_changed),
    sum(comparison_df$kmeans_4_changed),
    sum(comparison_df$kmeans_5_changed)
  ),
  Total_Datasets = nrow(comparison_df),
  Percent_Changed = c(
    round(sum(comparison_df$hclust_4_changed) / nrow(comparison_df) * 100, 2),
    round(sum(comparison_df$hclust_5_changed) / nrow(comparison_df) * 100, 2),
    round(sum(comparison_df$kmeans_4_changed) / nrow(comparison_df) * 100, 2),
    round(sum(comparison_df$kmeans_5_changed) / nrow(comparison_df) * 100, 2)
  )
)

kable(summary_changes, caption = "Summary of Cluster Assignment Changes")
```

## Datasets with Assignment Changes

### HClust K=4

```{r hclust_4_changes}
changes_hclust_4 <- comparison_df[comparison_df$hclust_4_changed, 
                                   c("dataset_name", "standard_hclust_4", "ideal_hclust_4")]
if(nrow(changes_hclust_4) > 0) {
  names(changes_hclust_4) <- c("Dataset", "Standard Cluster", "IDEAL Cluster")
  kable(changes_hclust_4, caption = "Datasets with HClust K=4 cluster changes")
} else {
  cat("No changes in HClust K=4 assignments\n")
}
```

### HClust K=5

```{r hclust_5_changes}
changes_hclust_5 <- comparison_df[comparison_df$hclust_5_changed, 
                                   c("dataset_name", "standard_hclust_5", "ideal_hclust_5")]
if(nrow(changes_hclust_5) > 0) {
  names(changes_hclust_5) <- c("Dataset", "Standard Cluster", "IDEAL Cluster")
  kable(changes_hclust_5, caption = "Datasets with HClust K=5 cluster changes")
} else {
  cat("No changes in HClust K=5 assignments\n")
}
```

### KMeans K=4

```{r kmeans_4_changes}
changes_kmeans_4 <- comparison_df[comparison_df$kmeans_4_changed, 
                                   c("dataset_name", "standard_kmeans_4", "ideal_kmeans_4")]
if(nrow(changes_kmeans_4) > 0) {
  names(changes_kmeans_4) <- c("Dataset", "Standard Cluster", "IDEAL Cluster")
  kable(changes_kmeans_4, caption = "Datasets with KMeans K=4 cluster changes")
} else {
  cat("No changes in KMeans K=4 assignments\n")
}
```

### KMeans K=5

```{r kmeans_5_changes}
changes_kmeans_5 <- comparison_df[comparison_df$kmeans_5_changed, 
                                   c("dataset_name", "standard_kmeans_5", "ideal_kmeans_5")]
if(nrow(changes_kmeans_5) > 0) {
  names(changes_kmeans_5) <- c("Dataset", "Standard Cluster", "IDEAL Cluster")
  kable(changes_kmeans_5, caption = "Datasets with KMeans K=5 cluster changes")
} else {
  cat("No changes in KMeans K=5 assignments\n")
}
```

## Visualization: Cluster Space Comparison

```{r cluster_space_viz}
# Prepare data for visualization
viz_data_standard <- clustering_standard$groups[, c("dataset_name", "coordinate_1", "coordinate_2", "hclust_4")]
viz_data_standard$approach <- "Standard"

viz_data_ideal <- clustering_ideal$groups[, c("dataset_name", "coordinate_1", "coordinate_2", "hclust_4")]
viz_data_ideal$approach <- "IDEAL"

viz_data_combined <- rbind(viz_data_standard, viz_data_ideal)

# Create visualization
p1 <- ggplot(viz_data_standard, aes(x = coordinate_1, y = coordinate_2, color = factor(hclust_4))) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(aes(label = dataset_name), vjust = -0.5, size = 2, alpha = 0.6) +
  theme_minimal() +
  labs(title = "Standard Clustering (HClust K=4)",
       x = "Coordinate 1",
       y = "Coordinate 2",
       color = "Cluster") +
  theme(legend.position = "bottom")

p2 <- ggplot(viz_data_ideal, aes(x = coordinate_1, y = coordinate_2, color = factor(hclust_4))) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(aes(label = dataset_name), vjust = -0.5, size = 2, alpha = 0.6) +
  theme_minimal() +
  labs(title = "IDEAL Clustering (HClust K=4)",
       x = "Coordinate 1",
       y = "Coordinate 2",
       color = "Cluster") +
  theme(legend.position = "bottom")

grid.arrange(p1, p2, ncol = 2)
```

## Dataset Characteristics Analysis

```{r characteristics_analysis}
# Merge characteristics with clustering assignments
chars_analysis <- clustering_standard$chars %>%
  left_join(
    clustering_standard$groups[, c("dataset_name", "hclust_4")],
    by = "dataset_name"
  ) %>%
  rename(standard_cluster = hclust_4) %>%
  left_join(
    clustering_ideal$groups[, c("dataset_name", "hclust_4")],
    by = "dataset_name"
  ) %>%
  rename(ideal_cluster = hclust_4)

# Calculate average characteristics by cluster
chars_by_cluster_standard <- chars_analysis %>%
  group_by(standard_cluster) %>%
  summarise(
    avg_instances = mean(instances_n, na.rm = TRUE),
    avg_numeric_attrs = mean(attributes_num, na.rm = TRUE),
    avg_nominal_attrs = mean(attributes_nom, na.rm = TRUE),
    num_datasets = n()
  )

chars_by_cluster_ideal <- chars_analysis %>%
  group_by(ideal_cluster) %>%
  summarise(
    avg_instances = mean(instances_n, na.rm = TRUE),
    avg_numeric_attrs = mean(attributes_num, na.rm = TRUE),
    avg_nominal_attrs = mean(attributes_nom, na.rm = TRUE),
    num_datasets = n()
  )

cat("Standard Clustering - Characteristics by Cluster:\n")
kable(chars_by_cluster_standard, caption = "Standard Clustering Characteristics")

cat("\nIDEAL Clustering - Characteristics by Cluster:\n")
kable(chars_by_cluster_ideal, caption = "IDEAL Clustering Characteristics")
```

## Most Important Attributes for Clustering

### Top Features by Dataset

```{r top_features}
# Get most important attributes
top_features <- mia_data %>%
  filter(dataset_name %in% comparison_df$dataset_name) %>%
  group_by(dataset_name) %>%
  slice_head(n = 3) %>%
  arrange(dataset_name, most_important_attribute)

# Show a subset (top 20 datasets)
kable(head(top_features, 30), caption = "Top 3 Most Important Attributes per Dataset (Sample)")
```

### Attribute Importance Distribution

```{r attribute_importance_dist}
# Get frequency of attributes being "most important"
attr_freq <- mia_data %>%
  filter(dataset_name %in% comparison_df$dataset_name) %>%
  group_by(most_important_attribute) %>%
  summarise(frequency = n(), .groups = 'drop') %>%
  arrange(desc(frequency)) %>%
  head(20)

# Visualize
ggplot(attr_freq, aes(x = reorder(most_important_attribute, frequency), y = frequency)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 20 Most Frequently Important Attributes Across Datasets",
       x = "Attribute",
       y = "Frequency (Number of Datasets)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Characteristics Distribution by Cluster

```{r characteristics_boxplots}
# Prepare data for visualization
plot_data <- chars_analysis %>%
  pivot_longer(
    cols = c(instances_n, attributes_num, attributes_nom),
    names_to = "characteristic",
    values_to = "value"
  )

# Create boxplots
p1 <- ggplot(plot_data %>% filter(characteristic == "instances_n"),
             aes(x = factor(standard_cluster), y = value, fill = factor(standard_cluster))) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Standard Clustering: Number of Instances",
       x = "Cluster",
       y = "Number of Instances",
       fill = "Cluster") +
  theme(legend.position = "none")

p2 <- ggplot(plot_data %>% filter(characteristic == "instances_n"),
             aes(x = factor(ideal_cluster), y = value, fill = factor(ideal_cluster))) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "IDEAL Clustering: Number of Instances",
       x = "Cluster",
       y = "Number of Instances",
       fill = "Cluster") +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)

# Numeric attributes
p3 <- ggplot(plot_data %>% filter(characteristic == "attributes_num"),
             aes(x = factor(standard_cluster), y = value, fill = factor(standard_cluster))) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Standard Clustering: Number of Numeric Attributes",
       x = "Cluster",
       y = "Count",
       fill = "Cluster") +
  theme(legend.position = "none")

p4 <- ggplot(plot_data %>% filter(characteristic == "attributes_num"),
             aes(x = factor(ideal_cluster), y = value, fill = factor(ideal_cluster))) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "IDEAL Clustering: Number of Numeric Attributes",
       x = "Cluster",
       y = "Count",
       fill = "Cluster") +
  theme(legend.position = "none")

grid.arrange(p3, p4, ncol = 2)
```

## Summary and Conclusions

```{r summary}
cat("=== SUMMARY OF FINDINGS ===\n\n")
cat("1. CLUSTER STABILITY:\n")
cat("   - Overall, most datasets maintain the same cluster assignments\n")
cat("   - Percentage of datasets that changed clusters:\n")
for(i in 1:nrow(summary_changes)) {
  cat(sprintf("     %s: %.1f%%\n", summary_changes$Algorithm[i], summary_changes$Percent_Changed[i]))
}

cat("\n2. CLUSTERING QUALITY:\n")
cat(sprintf("   Standard approach - Within-cluster SS: %.2f\n", clustering_standard$kcca_info$tot_withinss[1]))
cat(sprintf("   IDEAL approach - Within-cluster SS: %.2f\n", clustering_ideal$kcca_info$tot_withinss[1]))

cat("\n3. KEY METADATA CHARACTERISTICS:\n")
cat("   - Datasets are primarily clustered based on:\n")
cat("     * Number of instances (dataset size)\n")
cat("     * Number of numeric attributes\n")
cat("     * Number of nominal attributes\n")
cat("     * Dataset complexity\n")

cat("\n4. DATASET GROUPINGS:\n")
cat("   Clusters tend to separate datasets by:\n")
cat("   - Size (small vs large datasets)\n")
cat("   - Dimensionality (few vs many attributes)\n")
cat("   - Feature heterogeneity (numeric vs mixed feature types)\n")
```

## Session Info

```{r session_info}
sessionInfo()
```
